---
title: "Storm Data Analysis"
author: "Herizo RAJOELISON"
date: "1/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
sys_locale <- Sys.getlocale("LC_ALL")
Sys.setlocale("LC_ALL", "C")
```

# Analysys of Storm Data and the effects by type

Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

## Data Processing

Data is loaded from the initial source and then uncompressed inside the data folder. Some checks are made to ensure that the destination folder exists.

```{r, results='hide', message=FALSE}
library(dplyr)
library(R.utils)
stormdata_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
stormdata_filename <- "repdata_data_StormData.csv"
stormdata_filepath <- file.path("./data/", stormdata_filename)
stormdata_archivename <- "repdata_data_StormData.csv.bz2"
stormdata_archivepath <- file.path("./data/", stormdata_archivename)
if (!file.exists(stormdata_archivepath)) {
  download.file(stormdata_url, stormdata_archivepath)
}

bunzip2(stormdata_archivepath, stormdata_filepath, remove=FALSE, skip=TRUE)
```

The data is read inside the *stormdata* variable.

```{r, cache=TRUE}
stormdata <- read.csv(stormdata_filepath)

row_count <- nrow(stormdata)
col_count <- ncol(stormdata)
```

Before cleanup, the data contains `r col_count` columns and `r row_count` rows.

```{r}
column_labels <- colnames(stormdata)
print(column_labels)
```

```{r}
print(unique(trimws(stormdata$EVTYPE)))
```

## Results

There should be a section titled Results in which your results are presented.

```{r}
Sys.setlocale("LC_ALL", sys_locale)
```